---
title: "markdown test"
output: html_document
date: "2026-03-04"
---

Papaer title: Comprehensive single-cell transcriptomic and proteomic analysis reveals NK cell exhaustion and unique tumor cell evolutionary trajectory in non-keratinizing nasopharyngeal carcinoma

## Experimental Design

**Objective**: To investigate the role of NK cells and tumour cells in non-keratinising nasopharyngeal carcinoma (NK-NPC), which is closely associated with EBV infection.
非角化性鼻咽癌（NK-NPC）中 NK 细胞是否处于耗竭状态；

NK-NPC 中 肿瘤细胞是否存在特殊的演化轨迹，尤其是否与 EBV 感染状态有关。研究是围绕 NK 细胞功能和肿瘤细胞 evolutionary trajectory 展开的，而 NK-NPC 与 EBV 密切相关

**Method**: Collected three NK-NPC samples and three normal nasopharyngeal mucosa samples for proteomic analysis, combined with GSE162025 (NK-NPC single-cell transcriptome data) and GSE150825 (NLH single-cell transcriptome data) for analysis.

**Analytical steps:**

1. Analysis of proteomic data revealed a significant downregulation of proteins associated with NK cell-mediated cytotoxicity, suggesting that NK cell function may be suppressed. 


```{r eval=FALSE}
#install.packages('devtools')
#library(devtools)
#devtools::install_github('junjunlab/scRNAtoolVis')

# if not install ggunchull
#devtools::install_github("sajuukLyu/ggunchull", type = "source")

#BiocManager::install("CellChat")

# =========================
# 1. Load packages
# =========================

## Data import and pre processing
library(data.table)
library(readxl)
library(tibble)
library(dplyr)
library(stringr)

## Differential analysis
library(limma)

## Visualization
library(ggplot2)
library(ggrepel)
library(pheatmap)
library(patchwork)

## Functional enrichment
library(clusterProfiler)
library(enrichplot)
library(org.Hs.eg.db)
library(GSEABase)
library(msigdbr)

## Single-cell and downstream analyses
library(Seurat)
library(scRNAtoolVis)
library(CellChat)
library(copykat)
library(SCORPIUS)


```

```{r}
# proteomics
proteomics_file <- "re/rawdata/12967_2023_4112_MOESM1_ESM.xlsx"

data_pro <- read_xlsx(proteomics_file, sheet = "Proteins")

# Here we only keep:
#   1) "Gene Symbol" for protein-to-gene mapping
#   2) six sample-level normalized abundance columns
#      - F1, F3, F5: cancer samples
#      - F2, F4, F6: normal samples
#The goal is to build a protein expression matrix
# with rows = proteins (genes) and columns = 6 individual samples
# for downstream differential expression analysis


target_cols <- c(
  "Gene Symbol",
  "Abundances (Normalized): F1: Sample, cancer",
  "Abundances (Normalized): F3: Sample, cancer",
  "Abundances (Normalized): F5: Sample, cancer",
  "Abundances (Normalized): F2: Sample, normal",
  "Abundances (Normalized): F4: Sample, normal",
  "Abundances (Normalized): F6: Sample, normal"
)

data_pro$`gene symbol` <- str_split(data_pro$`gene symbol`,';',simplify = T)[,1]
data_pro <- data_pro[!duplicated(data_pro$`gene symbol`),]
data_pro <- data_pro[!is.na(data_pro$`gene symbol`),]
data_pro <- column_to_rownames(data_pro,'gene symbol')
data_pro <- data_pro[rowSums(!is.na(data_pro))==6,]

# deg
input_matrix <- log2(data_pro)
group <- as.factor(rep(c('NK_NPC','normal'),each=3))
design <- model.matrix(~0+group)
colnames(design) <- c('NK_NPC','normal')
rownames(design) <- colnames(input_matrix)
fit <- lmFit(input_matrix,design)
con.matrix <- makeContrasts('NK_NPC-normal',levels = design)
fit2 <- contrasts.fit(fit,contrasts = con.matrix)
fit2 <- eBayes(fit2)

# 差异筛选
options(digits = 4)
deg <- topTable(fit2,n=Inf)
up <- (deg$logFC>0.5)&(deg$P.Val<0.05)
down <- (deg$logFC<(-0.5))&(deg$P.Val<0.05)
change <- ifelse(up,'up',ifelse(down,'down','none'))
deg <- mutate(deg,change)
deg <- mutate(deg,'p' = -log10(deg$P.Val))
top5 <- rownames(top_n(deg[deg$change=='up',],5,logFC))
down5 <- rownames(top_n(deg[deg$change=='down',],-5,logFC))
deg$label <- ifelse(rownames(deg) %in% c(top5,down5),rownames(deg),NA)

# volcano
ggplot(deg,aes(x=logFC,y=p,color=change))+
  geom_point()+
  geom_vline(xintercept = 1,linetype=2,color='gray',linewidth=1)+
  geom_vline(xintercept = -1,linetype=2,color='gray',linewidth=1)+
  geom_hline(yintercept = -log10(0.05),linetype=2,color='gray',linewidth=1)+
  geom_text_repel(aes(logFC,label=label),
                  max.overlaps = 100000,size=3,segment.color='black',
                  box.padding=unit(1,'lines'))

# 差异蛋白pca
pca_data <- data_pro[rownames(deg[deg$change!='none',]),]
pca <- prcomp(t(log2(pca_data)),scale. = T)
xlab <- paste0('PCA_1(', round(summary(pca)$importance[2,1]*100,2),'%)')
ylab <- paste0('PCA_2(', round(summary(pca)$importance[2,2]*100,2),'%)')
ggplot(data.frame(pca$x),aes(x=PC1,y=PC2,color=group))+
  geom_point(size = 3)+stat_ellipse(level = 0.95, show.legend = F)+ 
  labs(x = xlab,y = ylab,title = 'PCA Analysis')+
  theme(plot.title = element_text(hjust = 0.5))+
  guides(color = guide_legend(title = 'Group'))
# heatmap
NK_cytotoxicity <- c('PRKCB','PTPN11','LCK','ITGAL','PIK3CB','MAPK3','CASP3')
pheatmap(data_pro[NK_cytotoxicity,],scale = 'row')
# KEGG
kegg_id <- bitr(rownames(deg[deg$change!='none',]),fromType = 'SYMBOL',toType = 'ENTREZID',OrgDb = org.Hs.eg.db)
kegg <- enrichKEGG(kegg_id$ENTREZID,'hsa',keyType = 'kegg')
kegg <- setReadable(kegg,OrgDb = org.Hs.eg.db,keyType = 'ENTREZID')
fc <- deg[deg$change!='none','logFC']
names(fc) <- rownames(deg[deg$change!='none',])
cnetplot(kegg,color.params = list(foldChange = fc,edge=T))


```
