---
title: "markdown test"
output: html_document
date: "2026-03-04"
---

Papaer title: Comprehensive single-cell transcriptomic and proteomic analysis reveals NK cell exhaustion and unique tumor cell evolutionary trajectory in non-keratinizing nasopharyngeal carcinoma

## Experimental Design

**Objective**: To investigate the role of NK cells and tumour cells in non-keratinising nasopharyngeal carcinoma (NK-NPC), which is closely associated with EBV infection.
非角化性鼻咽癌（NK-NPC）中 NK 细胞是否处于耗竭状态；

NK-NPC 中 肿瘤细胞是否存在特殊的演化轨迹，尤其是否与 EBV 感染状态有关。研究是围绕 NK 细胞功能和肿瘤细胞 evolutionary trajectory 展开的，而 NK-NPC 与 EBV 密切相关

**Method**: Collected three NK-NPC samples and three normal nasopharyngeal mucosa samples for proteomic analysis, combined with GSE162025 (NK-NPC single-cell transcriptome data) and GSE150825 (NLH single-cell transcriptome data) for analysis.

**Analytical steps:**

1. Analysis of proteomic data revealed a significant downregulation of proteins associated with NK cell-mediated cytotoxicity, suggesting that NK cell function may be suppressed. 


```{r eval=FALSE}
#install.packages('devtools')
#library(devtools)
#devtools::install_github('junjunlab/scRNAtoolVis')

# if not install ggunchull
#devtools::install_github("sajuukLyu/ggunchull", type = "source")

#BiocManager::install("CellChat")

# =========================
# 1. Load packages
# =========================
setwd("D:/re")
## Data import and pre processing
library(data.table)
library(readxl)
library(tibble)
library(dplyr)
library(stringr)

## Differential analysis
library(limma)

## Visualization
library(ggplot2)
library(ggrepel)
library(pheatmap)
library(patchwork)

## Functional enrichment
library(clusterProfiler)
library(enrichplot)
library(org.Hs.eg.db)
library(GSEABase)
library(msigdbr)

## Single-cell and downstream analyses
library(Seurat)
library(scRNAtoolVis)
library(CellChat)
library(copykat)
library(SCORPIUS)


```

```{r}
# proteomics
#setwd("D:/re")
# =========================
# 2. Prepare cleaning proteomic data
# =========================
proteomics_file <- "rawdata/12967_2023_4112_MOESM1_ESM.xlsx"

data_pro <- read_xlsx(proteomics_file, sheet = "Proteins")

# Here we only keep:
#   1) "Gene Symbol" for protein-to-gene mapping
#   2) six sample-level normalized abundance columns
#      - F1, F3, F5: cancer samples
#      - F2, F4, F6: normal samples
#The goal is to build a protein expression matrix
# with rows = proteins (genes) and columns = 6 individual samples
# for downstream differential expression analysis


target_cols <- c(
  "Gene Symbol",
  "Abundances (Normalized): F1: Sample, cancer",
  "Abundances (Normalized): F3: Sample, cancer",
  "Abundances (Normalized): F5: Sample, cancer",
  "Abundances (Normalized): F2: Sample, normal",
  "Abundances (Normalized): F4: Sample, normal",
  "Abundances (Normalized): F6: Sample, normal"
)

data_pro <- data_pro[, target_cols]
# take the first gene symbol from the colon and make it unique
data_pro$`Gene Symbol` <- str_split(data_pro$`Gene Symbol`, ';', simplify = T)[,1] 
# Remove leading/trailing spaces
data_pro$`Gene Symbol` <- trimws(data_pro$`Gene Symbol`)
data_pro <- data_pro[!is.na(data_pro$`Gene Symbol`) & data_pro$`Gene Symbol` != "", ]

data_pro <- data_pro[!duplicated(data_pro$`Gene Symbol`),]
# Use gene symbols as row names
data_pro <- column_to_rownames(data_pro,'Gene Symbol')
data_pro <- data_pro[rowSums(!is.na(data_pro))==ncol(data_pro),]
```


```{r}
# =========================
# 3. Differential expression analysis
# =========================
input_matrix <- as.matrix(data_pro)
storage.mode(input_matrix) <- "numeric"
input_matrix <- log2(input_matrix)

group <- factor(
  c("NK_NPC", "NK_NPC", "NK_NPC", "normal", "normal", "normal"),
  levels = c("NK_NPC", "normal")
)

design <- model.matrix(~ 0 + group)
colnames(design) <- c("NK_NPC", "normal")
rownames(design) <- colnames(input_matrix)

fit <- lmFit(input_matrix, design)
con.matrix <- makeContrasts(NK_NPC - normal, levels = design)
fit2 <- contrasts.fit(fit, contrasts = con.matrix)
fit2 <- eBayes(fit2, robust = TRUE)

deg <- topTable(fit2, n=Inf)

deg$change <- ifelse(
  deg$logFC > 0.5 & deg$adj.P.Val < 0.05, "up",
  ifelse(deg$logFC < -0.5 & deg$adj.P.Val < 0.05, "down", "none")
)
deg <- deg[order(deg$adj.P.Val, -abs(deg$logFC)), ]
deg <- rownames_to_column(deg, var = "Gene Symbol")

# value for volcano plot
deg$neglog10_fdr <- -log10(deg$adj.P.Val)

# split DEG results
deg_up <- deg[deg$change == "up", ]
deg_down <- deg[deg$change == "down", ]
deg_sig <- deg[deg$change != "none", ]

# check DEG counts
table(deg$change)
nrow(deg_up) # == 0
nrow(deg_down) # == 0
nrow(deg_sig) # == 0
### no proteins passing the predefined significance threshold after multiple-testing correction (|log2FC| > 0.5, FDR < 0.05) due to the small sample size 3*3 and large protein samples 1658.

deg$change <- ifelse(
  deg$logFC > 0.5 & deg$P.Value < 0.05, "up",
  ifelse(deg$logFC < -0.5 & deg$P.Value < 0.05, "down", "none")
)
deg <- deg[order(deg$P.Value, -abs(deg$logFC)), ]
deg <- rownames_to_column(deg, var = "Gene Symbol")

# value for volcano plot
deg$neglog10_fdr <- -log10(deg$P.Value)

# split DEG results
deg_up <- deg[deg$change == "up", ]
deg_down <- deg[deg$change == "down", ]
deg_sig <- deg[deg$change != "none", ]

# check DEG counts
table(deg$change)
nrow(deg_up) # == 0
nrow(deg_down) # == 0
nrow(deg_sig) # == 0


# =========================
# Volcano plot
# =========================


# select top 5 upregulated and top 5 downregulated proteins for labeling
top_up <- deg_up[order(deg_up$logFC, decreasing = TRUE), ]
top_down <- deg_down[order(deg_down$logFC, decreasing = FALSE), ]

top_up_genes <- head(top_up$`Gene Symbol`, 5)
top_down_genes <- head(top_down$`Gene Symbol`, 5)

deg$label <- ifelse(
  deg$`Gene Symbol` %in% c(top_up_genes, top_down_genes),
  deg$`Gene Symbol`,
  NA
)

lfc_cutoff <- 0.5
fdr_cutoff <- 0.05


p_volcano <- ggplot(deg, aes(x = logFC, y = neglog10_fdr, color = change)) +
  geom_point(size = 2, alpha = 0.8) +
  geom_vline(xintercept = c(-lfc_cutoff, lfc_cutoff), linetype = 2, color = "gray50") +
  geom_hline(yintercept = -log10(fdr_cutoff), linetype = 2, color = "gray50") +
  geom_text_repel(
    aes(label = label),
    na.rm = TRUE,
    size = 3,
    max.overlaps = 100
  ) +
  labs(
    x = "log2 fold change",
    y = "-log10 P value",
    color = "Change",
    title = "Proteomics volcano plot"
  ) +
  theme_classic()

p_volcano

```


```{r}
# pca
pca_data <- data_pro[rownames(deg[deg$change!='none',]),]
pca <- prcomp(t(log2(pca_data)),scale. = T)
xlab <- paste0('PCA_1(', round(summary(pca)$importance[2,1]*100,2),'%)')
ylab <- paste0('PCA_2(', round(summary(pca)$importance[2,2]*100,2),'%)')
ggplot(data.frame(pca$x),aes(x=PC1,y=PC2,color=group))+
  geom_point(size = 3)+stat_ellipse(level = 0.95, show.legend = F)+ 
  labs(x = xlab,y = ylab,title = 'PCA Analysis')+
  theme(plot.title = element_text(hjust = 0.5))+
  guides(color = guide_legend(title = 'Group'))
# heatmap
NK_cytotoxicity <- c('PRKCB','PTPN11','LCK','ITGAL','PIK3CB','MAPK3','CASP3')
pheatmap(data_pro[NK_cytotoxicity,],scale = 'row')
# KEGG
kegg_id <- bitr(rownames(deg[deg$change!='none',]),fromType = 'SYMBOL',toType = 'ENTREZID',OrgDb = org.Hs.eg.db)
kegg <- enrichKEGG(kegg_id$ENTREZID,'hsa',keyType = 'kegg')
kegg <- setReadable(kegg,OrgDb = org.Hs.eg.db,keyType = 'ENTREZID')
fc <- deg[deg$change!='none','logFC']
names(fc) <- rownames(deg[deg$change!='none',])
cnetplot(kegg,color.params = list(foldChange = fc,edge=T))


```
